AppPAL for Android

### 2015-10-02 paper.tex
further updates and some tweaks to the bibliography
Final version :-S

% vim: set nowrap tw=0:
\documentclass[]{llncs}

@@

\title{AppPAL for Android}
\subtitle{Capturing and Checking Mobile App Policies}
% \numberofauthors{2}
\author{Joseph Hallett \and David Aspinall }
\institute{University of Edinburgh}

@@

\begin{abstract}
  It can be difficult to find mobile apps that respect your security and privacy.
  Employers rely on employees enforcing company policies about apps at the workplace correctly.
  Users must judge apps by the information shown to them by the store.
  Only 17\% of users pay attention to an apps permissions during installation~\cite{Felt:2012hm} and
  most users do not understand how permissions relate to the capabilities of an app~\cite{Kelley:2012bw}.
  We present AppPAL: a policy language for Android to describe app installation  policies.
  AppPAL goes beyond existing policy enforcement tools, like Kirin~\cite{Enck:2008vn}, adding delegation relationships and the connecting to static analysis tools.
  We use AppPAL to explore what kinds of policies are used when installing apps, demonstrating the privacy paradox when comparing user's privacy views to their actual behaviour.
\end{abstract}

\section{Introduction \comment{1 page}}
\label{sec:introduction}

Finding the right apps can be tricky.
Users need to discover which are not going to abuse their data.
This can be difficult as it isn't obvious how apps use the data each has access to.
Consider a user attempting to buy a flashlight app.
By searching the Play store the user is presented with a long list of apps.
Clicking through each one they can find the permissions each requests but not the reasons why each was needed.
They can see review scores from users but not experts who can run tools to check apps for problems and issues like SSL misconfigurations~\cite{Fahl:2012dj}.
If they want to use the app at work will it break their employers rules for mobile usage?

App stores give some information about their apps; descriptions of the app and screenshots as well as review scores.
Android apps show a list of permissions when they're first installed.
Soon new apps will display permissions requests when the app first tries to access sensitive data (such as contacts or location information).
Users do not understand how permissions relate to their device~\cite{Felt:2012hm,Thompson:2013eb}.
Ultimately the decision which apps to use and which permissions to grant must be made by the device user.

@@

These policies can only be enforced by the users continuously making the correct decision when prompted about apps.
This is error-prone.
An alternative is to write the policy down and make the computer enforce it.
To implement this we use a logic of authorization---a language designed to express rules about permissible actions.
The policy is written in the logic and enforced by checking the policy is satisfied.

We present AppPAL, an instantiation of Becker~\etal's SecPAL~\cite{Becker:2006vh} with constraints\footnote{A constraint makes use of information checkable using information external to the language, such as the time of day or output of a static analysis tool.} and predicates that allow us to decide which apps to run or install.
The language allows us to reason about apps using statements from third parties.
The implementation allows us to enforce the policies on a device.
We can express trust relationships amongst these parties; use constraints to do additional checks, such as using static analysis results.
This lets us enforce more complex policies than existing tools such as Kirin~\cite{Enck:2009ko} which are limited to permissions checks.

@@

\noindent
\begin{minipage}{\linewidth}
  \begin{minipage}{0.5\linewidth}
    \begin{lstlisting}
"alice" says App isRunnable
  if "home-policy" isMetBy(App)
  where at("work") = false.
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{0.5\linewidth}
    \begin{lstlisting}
"alice" says App isRunnable
  if "work-policy" isMetBy(App)
  where beforeHourOfDay("17") = true.
    \end{lstlisting}
  \end{minipage}
\end{minipage}

We can delegate policy specification to third parties or roles, and assign principals to roles:
\begin{lstlisting}
"alice" says "it-department" can-say "work-policy" isMetBy(App).
"alice" says "alice" can-act-as "it-department".
\end{lstlisting}
We can write policies specifying which permissions an app must or must not have by its app store categorization.
For example it would be okay allowing a photography app access to the camera, but not to the location data if the user doesn't want their photos geotagged.
\begin{lstlisting}
"alice" says App isRunnable
  if "permissions-policy" isMetBy(App).
"alice" says "permissions-policy" isMetBy(App)
  if App isAnApp
  where
    category(App, "Photography"),
    hasPermission(App, "LOCATION") = false,
    hasPermission(App, "CAMERA") = true.
\end{lstlisting}

@@

AppPAL can act as a \emph{``glue"} between static analysis tools and the app installation policies device owner's are trying to enforce.
This avoids creating tools with hard-coded fixed policies
For example a store might not want to sell apps with SSL errors or malicious apps as determined by their anti-virus.
Using AppPAL we can combine tools for checking apps to implement the store's policies.
\begin{lstlisting}
"play-store" says App isInstallable
  if App isAnApp
  where mallodroidCheck(App) = true,
         mcafeeAVCheck(App) = true.
\end{lstlisting}

% \noindent
% In this paper we have:
% \begin{itemize}
%   \item
%     Described a scenario where an employer has a policy they want to enforce for their employees~(\autoref{sec:problem});
%     and shown how the employer's policy could be implemented using AppPAL~(\autoref{sec:idea}) and installed on Alice's phone.
%   \item
%     Described our implementation of the AppPAL language as a library for Android and the JVM.
%     We show how the language can describe app properties and Android security policies (\autoref{ssec:language}) as well as how to check policies (\autoref{ssec:eval}).
%   \item
%     Shown the need for policy tools by demonstrating the privacy paradox holds for user privacy policies with app installations (\autoref{sec:demonstation}).
% \end{itemize}

\section{Enforcing a policy at work}
\label{sec:problem}

@@

\section{Expressing policies in AppPAL}
\label{sec:idea}

In \autoref{sec:problem} Alice and Emma had policies they wanted to enforce but no means to do so.
Instead of using several different tools to enforce Emma's policy disjointedly, we could use an authorization logic.
In \autoref{lst:corporate} we give an AppPAL policy implementing Emma's app concerns on Alice's phone.
% A pictorial representation, is given in \autoref{fig:emmas_policy}; color splits statements by each speaker.

@@

"emma" says "google-play" can-say
  App isBuyable. @\label{lst:corporate_4}@
    \end{lstlisting}
  \end{minipage}\begin{minipage}[t]{0.5\textwidth}
    \begin{lstlisting}[numbers=left,
                       escapeinside={@}{@},
                       firstnumber=15,
                       basicstyle=\scriptsize\ttfamily,
                       stringstyle=\scriptsize\sffamily,
                       keywordstyle=\scriptsize\slshape]
"emma" says "anti-virus-policy" isMetBy(App) @\label{lst:corporate_5}@
  if App isAnApp
  where
    mcAfeeVirusCheck(App) = false.

@@

"emma" says
  "no-location-permissions" isMetBy(App) @\label{lst:corporate_7}@
    if App isAnApp
    where
      hasPermission(App, "LOCATION")=false.
\end{lstlisting}
\end{minipage}
\caption{AppPAL policy implementing Emma's security requirements.}
\label{lst:corporate}
\end{figure}

In \autoref{lst:corporate_1} Alice gives Emma the ability to specify whether an \code{App} (a variable) \code{isRunnable} (a predicate).
She allows her to delegate the decision further if she chooses (\code{can-say inf}).
Next in \autoref{lst:corporate_2} Emma specifies her concerns as policies to be met (the \code{isMetBy()} predicate that takes an app as its argument).
If Emma can be convinced all these policies are met then she will say the \code{App isRunnable}.
In \autoref{lst:corporate_3} and \autoref{lst:corporate_4} Emma specifies that an app meets the \code{reputable-policy} if the \code{App isBuyable};
  with \code{"google-play"} specified as the decider of what is buyable or not.
Google is not allowed to delegate the decision further (\code{can-say} as oposed to \code{can-say inf}).
In other words Google is not allowed to specify Amazon as a supplier of apps as well.
Google must say what is buyable directly for Emma to accept it.
Emma specifies the \code{"anti-virus-policy"} in \autoref{lst:corporate_5}.
Here we use a constraint.
When checking the policy the \code{mcAfeeVirusCheck} should be run on the \code{App}.
Only if this returns \code{false} will the policy be met.
To specify the \code{"no-tracking-policy"} Emma says that the \code{"no-location-permissions"} rules implement the \code{"no-tracking-policy"} (\autoref{lst:corporate_6}).
Emma specifies this in \autoref{lst:corporate_7} by checking the app is missing two permissions.

Alice wants to install a new app (\code{com.facebook.katana}) on her phone.
To meet Emma's policy the AppPAL checker needs to collect statements to show the app meets the \code{isRunnable} predicate.
Specifically it needs:
\begin{itemize}
  \item\code{"emma" says "com.facebook.katana" isAnApp.}
    A simple typing statement that can be generated for all apps as they are encountered.
    This helps keep the number of assertions in the policy low aiding readability.
  \item\code{"google-play" says "com.facebook.katana" isReputable.}
    Required to convince Emma the app came from a reputable source.
    It should be able to obtain this statement from the Play store as the app is available there.
  \item\code{"emma" says "anti-virus-policy" isMetBy("com.facebook.katana").}
    She can obtain this by running the \ac{av} program on her app.
  \item\code{"emma" says "no-locations-permissions" isMetBy("com.facebook.katana").}
    Needed to show the App meets Emma's no-tracking-policy.
    Emma will say this if after examining the app the location permissions are missing.
\end{itemize}
These last two statements require the checker to do some extra checks to satisfy the constraints.
To get the third statement it must run the \ac{av} program on her app and check the result.
The results from the \ac{av} program may change with time as it's signatures are updated;
  so the checker must re-run this check every time it wants to obtain the statement connected to the constraint.
For the forth statement the checker needs to check the permissions of the app.
It could do this by looking in the \texttt{MANIFEST.xml} inside the app itself, or through the Android package manager if it is running on a device.

@@

\begin{figure}
  \newcommand{\bracetext}[1]{\text{\sffamily #1}}
  \newcommand{\smalltext}[1]{\text{\ttfamily\scriptsize #1}}
  \centering
  \begin{minipage}{0.49\linewidth}
    \begin{equation*}\small
      \begin{array}{r l}\footnotesize
        \overbrace{\smalltext{`user'}}^{\bracetext{speaker}} &
        \smalltext{ says }\overbrace{\overbrace{\smalltext{ App }}^{\bracetext{subject}}\overbrace{\smalltext{ isRunnable}}^{\bracetext{predicate}}}^{\bracetext{fact}} \\
        & \overbrace{\smalltext{ if App isFree}}^{\bracetext{condition}} \\
        & \overbrace{\smalltext{ where hasPermission(App, `INTERNET') = true}}^{\bracetext{constraint}}.
      \end{array}
    \end{equation*}
  \end{minipage}
  \begin{minipage}{0.49\linewidth}
  \newcommand{\nonterminal}[1]{$\langle$#1$\rangle$}
  \newcommand{\terminal}[1]{\textbf{#1}}
  \begin{tabular}{r c l}
    \footnotesize
    \nonterminal{Assertion} & $\coloneqq$ & \nonterminal{E} \terminal{says} \nonterminal{Fact} \\
                            &             & \hspace{1em}(\terminal{if} (\nonterminal{Fact}\terminal{,})+)?\terminal{.} \\
                            &             & \hspace{1em}(\terminal{where} \nonterminal{Constraint})? \\
    \nonterminal{Fact}      & $\coloneqq$ & \nonterminal{E} (\terminal{isRunable} $\vert$ $\ldots$) \\
                            & $\vert$     & \nonterminal{E} \terminal{can-say} \terminal{inf}? \nonterminal{Fact} \\
                            & $\vert$     & \nonterminal{E} \terminal{can-act-as} \nonterminal{E} \\
    \nonterminal{E}         & $\coloneqq$ & \terminal{Variable} $\vert$ \terminal{`constant'}
  \end{tabular}
  \end{minipage}
  \caption{Structure and simplified grammar of an AppPAL assertion.}
  \label{fig:assertion}
\end{figure}

@@

Splitting the decision about whether an app is runnable into a series of policies that must be met gives us flexibility in how the decision is made.
It allows us to describe multiple means of making the same decision, and provide backup routes when one fails.
Some static analysis tools are not quick to run.
Even taking minutes to run a battery draining analysis can be undesirable.
If a user wants to download an app quickly they may not be willing to wait to check that a policy is met.

In \autoref{sec:problem} and \autoref{sec:idea} we described a \emph{no-tracking-policy} to prevent a user's location being leaked.
In Emma's policy we checked this using the app's permissions.
If the app couldn't get access to the GPS sensors (using the permissions) then it met this policy.
Some apps may want to access this data, but may not leak it.
We could use a taint analysis tool to detect this (e.g.~FlowDroid~\cite{Fritz:2013vi}).
Our policy now becomes:

@@

"emma" says "no-locations-permissions" isMetBy(App)
  if App isAnApp
  where
    hasPermission(App, "ACCESS_FINE_LOCATION") = false,
    hasPermission(App, "ACCESS_COARSE_LOCATION") = false.

@@

"emma" says "location-taint-analysis" isMetBy(App)
  if App isAnApp
  where
    flowDroidCheck(App, "Location", "Internet") = false.
\end{lstlisting}

@@

\begin{lstlisting}
"emma" says "relaxed-no-tracking-policy" canActAs "no-tracking-policy".
"emma" says "relaxed-no-tracking-policy" isMetBy(App)
  if App hasCategory("tracking")
  where
    mallodroidSSLCheck(App) = false,
    connectionsCheck(App, "[https://emma.com]") = true.
\end{lstlisting}

@@

\subsection{Policy checking}
\label{ssec:eval}

AppPAL has the same policy checking rules as SecPAL~\cite{Becker:2006vh}.
AppPAL uses an assertion context of known facts and rules, as well as facts deduced while checking.
We do not use Becker~\etal's DatalogC based checking algorithm, as no DatalogC library exists for Android.
We have implemented the rules directly in Java.
Pseudo-code is given in \autoref{fig:pseudocode}.

On a mobile device memory is at a premium.
We would like to keep the assertion context as small as possible.
For some assertions (like \code{isAnApp}) we derive them by checking the arguments at evaluation time.
This gives us greater control of the evaluation and how the assertion context is created.
For example, when checking the \code{isAnApp} predicate;
  we can fetch the assertion that the subject is an app based on the app in question.

@@

We envisage AppPAL running on a mobile phone, checking apps when they are installed.
Since policy checks may involve inspecting many rules and constraints one may ask whether the checking will be acceptably fast.
Downloading and installing an app takes about 30 seconds on a typical Android phone over wifi.
If checking a policy delays this even further a user may become annoyed and disable AppPAL.

The policy checking procedure is at its slowest when having to delegate repeatedly;
   the depth of the delegation tree is the biggest factor for slowing the search.
Synthetic benchmarks were created to check that the checking procedure performed acceptably.
Each benchmark consisted of a chain of delegations.
The \emph{1 to 1} benchmark consists of a repeated delegation between all of the principals.
In the \emph{1 to 2} benchmark each principal delegated to 2 others and in the \emph{1 to 3} benchmark each principal delegated to 3 others.
These benchmarks are reasonable as they model the slowest kinds of policies to evaluate---though worse ones could be designed by delegating even more.

@@

Throughout we have asserted that user's have policies and that there is a need for policy enforcement tools.
Corporate mobile security \ac{byod} policies have started appearing and NIST have issued recommendations for writing them~\cite{Scarfone:2009vy,Souppaya:2013jf}.
In a study of 725 Android users, Lin~\etal~found four patterns that characterise user privacy preferences for apps~\cite{Sadeh:2014vq}.
Using app installation data from Carat~\cite{Oliner:2013ht} we used AppPAL to find the apps satisfying each policy Lin~\etal~identified and measured the extent each user was following them.

Lin~\etal~identified four types of user.
The \emph{Conservative} (C) users were uncomfortable allowing an app access to any personal data for any reason.
The \emph{Unconcerned} (U) users felt okay allowing access to most data for almost any reason.
The \emph{Advanced} (A) users were comfortable allowing apps access to location data but not if it was for advertising reasons.
Opinions in the largest cluster, \emph{Fencesitters} (F), varied but were broadly against collection of personal data for advertising purposes.
We wrote AppPAL policies to describe each of these behaviours as increasing sets of permissions.
These simplify the privacy policies identified as by Lin~\etal~as we do not take into account the reason each app might have been collecting each permission.
Using AppPAL we could write more precise rules if we could determine why each permission was requested.
Lin~\etal~used Androguard~\cite{Desnos:ub} as well as manual analysis to determine the precise reasons for each permission~\cite{Sadeh:2014vq}.

@@

McAfee classify malware into several categories.
The \emph{malicious} and \emph{trojan} categories describe traditional malware.
Other categories classify \ac{pup} such as aggressive adware.
Using AppPAL we can write policies to differentiate between different kinds of malware, characterising users who allow dangerous apps and those who install poor quality ones.
\begin{lstlisting}
"user" says "mcafee" can-say
  "malware" isKindOf(App).
"mcafee" says "trojan" can-act-as "malware".
"mcafee" says "pup" can-act-as "malware".
\end{lstlisting}
If a user is enforcing a privacy policy we might also expect them to install less malware.
We can check this by using AppPAL policies to measure the number of malwares each user had installed.

We want to test how well policies capture user behaviour.
Installation data was taken from a partially anonymized\footnote{Users are replaced with incrementing numbers, app names are replaced with hashes to protect sensitive names.} database of installed apps captured by Carat~\cite{Oliner:2013ht}.
By calculating the hashes of known package names we see who installed what.
The initial database has over 90,000 apps and 55,000 users.
On average each user installed around 90 apps each; 4,300 apps have known names.
Disregarding system apps (such as \texttt{com.android.vending}) and very common apps (Facebook, Dropbox, Whatsapp, and Twitter) we reduced the set to an average of 20 known apps per user.
To see some variations in app type, we considered only the 44,000 users who had more than 20 known apps.
Using this data, and the apps themselves taken from the Google Play Store and Android Observatory~\cite{Barrera:2012iba}, we checked which apps satisfied which policies.

\begin{figure}\centering
  \subfigure[Use of policies modelling user behaviour.]{%
    \includegraphics[width=0.48\linewidth]{./figures/lin.eps}
    \label{fig:lin}}
  \subfigure[Percentage of malware in installed apps for users installing some malicious apps.]{%
    \includegraphics[width=0.48\linewidth]{./figures/malware.eps}
    \label{fig:malware}}
  \caption{Policy compliance graphs.}
\end{figure}

>>>>

% vim: set nowrap tw=0:
\documentclass[]{llncs}
\usepackage{microtype}

@@

\title{AppPAL for Android}
\subtitle{Capturing and Checking Mobile App Policies}
\titlerunning{AppPAL for Android}
% \numberofauthors{2}
\author{Joseph Hallett \and David Aspinall }
\institute{School of Informatics, University of Edinburgh}

@@

\begin{abstract}
  It can be difficult to find mobile apps that respect your security and privacy.
  Businesses rely on employees enforcing company mobile device policies correctly.
  Users must judge apps by the information shown to them by the store.
  Only 17\% of users pay attention to an apps permissions during installation~\cite{Felt:2012hm} and most users do not understand how permissions relate to the capabilities of an app~\cite{Kelley:2012bw}.
  To address these problems, we present AppPAL: a machine-readable policy language for Android that describes precisely when apps are acceptable.
  AppPAL goes beyond existing policy enforcement tools, like Kirin~\cite{Enck:2008vn}, adding delegation relationships to allow a variety of authorities to contribute to a decision.
  AppPAL also acts as a ``glue'', allowing connection to a variety of local constraint checkers (e.g., static analysis tools, packager manager checks) to combine their results.
  As well as introducing AppPAL and some examples, we apply it to explore whether real users follow certain intended policies in practice, finding privacy preferences and actual behaviour are not always aligned, in the absence of a rigorous enforcement mechanism.
  \end{abstract}

\section{Introduction}
\label{sec:introduction}

Finding the right apps can be tricky.
Users need to discover which are not going to abuse their data.
This can be difficult as it isn't obvious how apps use the data each has access to.
Consider a user attempting to buy a flashlight app.
By searching the Play store the user is presented with a long list of apps.
Clicking through each one they can find the permissions each requests but not the reasons why each was needed.
They can see review scores from users but from tools to check apps for problems and issues like SSL misconfigurations~\cite{Fahl:2012dj}.
If they want to use the app at work will it break their employers rules for mobile usage?

App stores give some information about their apps; descriptions, screenshots and review scores.
Android apps show a list of permissions when they're first installed.
Soon new apps will display permissions requests when the app first tries to access sensitive data (such as contacts or location information).
Users do not understand how permissions relate to their device~\cite{Felt:2012hm,Thompson:2013eb}.
Ultimately the decision which apps to use and which permissions to grant must be made by the device user.

@@

These policies can only be enforced by the users continuously making the correct decision when prompted about apps.
An alternative is to write the policy down and make the computer enforce it.
To implement this we use a logic of authorization---a language designed to express rules about permissible actions.
The policy is written in the logic and enforced by checking the policy is satisfied.

We present AppPAL, an instantiation of Becker~\etal's SecPAL~\cite{Becker:2006vh} with constraints\footnote{A constraint makes use of information checkable using information external to the language, such as the time of day or output of a static analysis tool.} and predicates that allow us to decide which apps to run or install.
The language allows us to reason about apps using statements from third parties.
AppPAL allows us to enforce the policies on a device.
We can express trust relationships amongst these parties; use constraints to do additional checks, such as using static analysis results.
This lets us enforce more complex policies than existing tools such as Kirin~\cite{Enck:2009ko} which are limited to permissions checks.

@@

\noindent
\begin{minipage}{\linewidth}
  \begin{minipage}{0.5\linewidth}
    \begin{lstlisting}
"alice" says App isRunnable
  if "home-policy" isMetBy(App)
  where at("work") = False.
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{0.5\linewidth}
    \begin{lstlisting}
"alice" says App isRunnable
  if "work-policy" isMetBy(App)
  where beforeHourOfDay("17") = True.
    \end{lstlisting}
  \end{minipage}
\end{minipage}

We can delegate policy specification to third parties or roles, and assign principals to roles:
\begin{lstlisting}
"alice" says "it-department" can-say "work-policy" isMetBy(App).
"alice" says "alice" can-act-as "it-department".
\end{lstlisting}
We can write policies specifying which permissions an app must or must not have by its app store categorization.
For example it would be okay allowing a photography app access to the camera, but not to location data if the user doesn't want their photos geotagged.
\begin{lstlisting}
"alice" says App isRunnable
  if "permissions-policy" isMetBy(App).
"alice" says "permissions-policy" isMetBy(App)
  if App isAnApp
  where
    category(App, "Photography"),
    hasPermission(App, "LOCATION") = False,
    hasPermission(App, "CAMERA") = True.
\end{lstlisting}

@@

AppPAL can act as a \emph{``glue"} between static analysis tools and the app installation policies device owners are trying to enforce.
This avoids creating tools with hard-coded fixed policies
For example a store might not want to sell apps with SSL errors or malicious apps as determined by their anti-virus.
Using AppPAL we can combine tools for checking apps to implement the store's policies.
\begin{lstlisting}
"play-store" says App isSellable
  if App isAnApp
  where mallodroidCheck(App) = True,
         mcafeeAVCheck(App) = True.
\end{lstlisting}

\section{Enforcing A Policy At Work}
\label{sec:problem}

@@

\section{Expressing Policies In AppPAL}
\label{sec:idea}

In \autoref{sec:problem} Alice and Emma had policies they wanted to enforce but no means to do so.
Instead of using several tools to enforce Emma's policy disjointedly, we could use an authorization logic.
In \autoref{lst:corporate} we give an AppPAL policy implementing Emma's app concerns on Alice's phone.
% A pictorial representation, is given in \autoref{fig:emmas_policy}; color splits statements by each speaker.

@@

"emma" says "google-play" can-say
  App isBuyable. @\label{lst:corporate_4}@
    \end{lstlisting}
  \end{minipage}\begin{minipage}[t]{0.5\textwidth}
    \begin{lstlisting}[numbers=left,
                       escapeinside={@}{@},
                       firstnumber=15,
                       basicstyle=\scriptsize\ttfamily,
                       stringstyle=\scriptsize\sffamily,
                       keywordstyle=\scriptsize\slshape]
"emma" says "anti-virus-policy" isMetBy(App) @\label{lst:corporate_5}@
  if App isAnApp
  where
    mcAfeeVirusCheck(App) = False.

@@

"emma" says
  "no-location-permissions" isMetBy(App) @\label{lst:corporate_7}@
    if App isAnApp
    where
      hasPermission(App, "LOCATION")=False.
\end{lstlisting}
\end{minipage}
\caption{AppPAL policy implementing Emma's security requirements.}
\label{lst:corporate}
\end{figure}

In \autoref{lst:corporate_1} Alice less Emma  specify whether an \code{App} (a variable) \code{isRunnable};
she allows her to delegate the decision further (\code{can-say}\/~\code{inf}).
Emma specifies her concerns as policies to be met in \autoref{lst:corporate_2}:
  if Emma is convinced all these are met then she will say the \code{App isRunnable}.
In \autoref{lst:corporate_3} and \autoref{lst:corporate_4} Emma specifies that an app meets the \code{reputable-policy} if the \code{App isBuyable};
  with \code{"google-play"} as the decider of what is buyable or not.
Google is not allowed to delegate the decision further, i.e.~Google is not allowed to specify Amazon as a supplier of apps as well.
Emma specifies the \code{"anti-virus-policy"} in \autoref{lst:corporate_5} using a constraint.
When checking the policy the \code{mcAfeeVirusCheck} should be run on the \code{App}.
Only if this returns \code{false} will the policy be met.
To specify the \code{"no-tracking-policy"} Emma says that the \code{"no-location-permissions"} rules implement the \code{"no-tracking-policy"} (\autoref{lst:corporate_6}).
Emma specifies this in \autoref{lst:corporate_7} by checking the app is missing two permissions.

Alice wants to install a new app (\code{com.facebook.katana}) on her phone
she needs to collect statements to show the app meets the \code{isRunnable} predicate.
Namely:
\begin{itemize}
  \item\code{"google-play" says "com.facebook.katana" isReputable.}
    Required to convince Emma the app came from a reputable source.
  \item\code{"emma" says "anti-virus-policy" isMetBy("com.facebook.katana").}
    She can obtain this by running the \ac{av} program on her app.
  \item\code{"emma" says "no-locations-permissions" isMetBy("com.facebook.katana").}
    Needed to show the App meets Emma's no-tracking-policy.
    Emma will say this if after examining the app the location permissions are missing.
\end{itemize}
These last two statements require the checker to do some extra checks to satisfy the constraints.
To get the second statement AppPAL must run the \ac{av} program on her app and check the result.
The results from the \ac{av} program may change with time as it's signatures are updated;
  so the checker must re-run this check every time it wants to obtain the statement connected to the constraint.
For the third statement the AppPAL checker needs to examine the permissions of the app.
It could do this by looking in the \texttt{MANIFEST.xml} inside the app itself, or through the Android package manager if it is running on a device.

@@

\begin{figure}
  \newcommand{\bracetext}[1]{\text{\sffamily #1}}
  \newcommand{\smalltext}[1]{\text{\ttfamily\scriptsize #1}}
  \centering
  \begin{minipage}{0.49\linewidth}
    \begin{equation*}\small
      \begin{array}{r l}\footnotesize
        \overbrace{\smalltext{`user'}}^{\bracetext{speaker}} &
        \smalltext{ says }\overbrace{\overbrace{\smalltext{ App }}^{\bracetext{subject}}\overbrace{\smalltext{ isRunnable}}^{\bracetext{predicate}}}^{\bracetext{fact}} \\
        & \overbrace{\smalltext{ if App isFree}}^{\bracetext{condition}} \\
        & \overbrace{\smalltext{ where hasPermission(App, `INTERNET') = True}}^{\bracetext{constraint}}.
      \end{array}
    \end{equation*}
  \end{minipage}
  \begin{minipage}{0.49\linewidth}
  \newcommand{\nonterminal}[1]{$\langle$#1$\rangle$}
  \newcommand{\terminal}[1]{\textbf{#1}}
  \begin{tabular}{r c l}
    \footnotesize
    \nonterminal{Assertion} & $\coloneqq$ & \nonterminal{E} \terminal{says} \nonterminal{Fact} \\
                            &             & \hspace{1em}(\terminal{if} (\nonterminal{Fact}\terminal{,})+)?\terminal{.} \\
                            &             & \hspace{1em}(\terminal{where} \nonterminal{Constraint})? \\
    \nonterminal{Fact}      & $\coloneqq$ & \nonterminal{E} (\terminal{isRunable} $\vert$ $\ldots$) \\
                            & $\vert$     & \nonterminal{E} \terminal{can-say} \terminal{inf}? \nonterminal{Fact} \\
                            & $\vert$     & \nonterminal{E} \terminal{can-act-as} \nonterminal{E} \\
    \nonterminal{E}         & $\coloneqq$ & \terminal{Variable} $\vert$ \terminal{`constant'}
  \end{tabular}
  \end{minipage}
  \caption{Structure and simplified grammar of an AppPAL assertion.}
  \label{fig:assertion}
\end{figure}

@@

Splitting the decision about whether an app is runnable into a series of policies that must be met gives us flexibility in how the decision is made.
It allows us to describe multiple means of making the same decision, and provide backup routes when one fails.
Some static analysis tools are not quick to run.
Even taking minutes to run a battery draining analysis can be undesirable:
if a user wants to download an app quickly they may not be willing to wait to check that a policy is met.

In \autoref{sec:problem} and \autoref{sec:idea} we described a \emph{no-tracking-policy} to prevent a user's location being leaked.
In Emma's policy we checked this using the app's permissions;
if the app couldn't get access to the GPS sensors (using the permissions) then it met this policy.
Some apps may want to access this data, but may not leak it.
We could use a taint analysis tool to detect this (e.g.~FlowDroid~\cite{Fritz:2013vi}).
Our policy now becomes:

@@

"emma" says "no-locations-permissions" isMetBy(App)
  if App isAnApp
  where
    hasPermission(App, "ACCESS_FINE_LOCATION") = False,
    hasPermission(App, "ACCESS_COARSE_LOCATION") = False.

@@

"emma" says "location-taint-analysis" isMetBy(App)
  if App isAnApp
  where
    flowDroidCheck(App, "Location", "Internet") = False.
\end{lstlisting}

@@

\begin{lstlisting}
"emma" says "relaxed-no-tracking-policy" canActAs "no-tracking-policy".
"emma" says "relaxed-no-tracking-policy" isMetBy(App)
  if App hasCategory("tracking")
  where
    mallodroidSSLCheck(App) = False,
    connectionsCheck(App, "[https://emma.com]") = True.
\end{lstlisting}

@@

\subsection{Policy Checking}
\label{ssec:eval}

AppPAL has the same policy checking rules as SecPAL~\cite{Becker:2006vh}.
AppPAL uses an assertion context of known facts and rules, as well as facts deduced while checking.
While Becker~\etal used a DatalogC based checking algorithm, we have implemented the rules directly in Java as no DatalogC library is currently available for Android.
Pseudo-code is shown in \autoref{fig:pseudocode}.

On a mobile device memory is at a premium.
We would like to keep the assertion context as small as possible.
For some assertions (like \code{isAnApp}) we derive them by checking the arguments at evaluation time.
This gives us greater control of the evaluation and how the assertion context is created.
For example, when checking the \code{isAnApp} predicate;
  we can fetch the assertion that the subject is an app based on the app in question.
When delegating we will also be able to request facts from the delegated party dynamically; though implementing this is future work.

@@

When AppPAL runs on a mobile phone apps should be checked as they are installed.
Since policy checks may involve inspecting many rules and constraints one may ask whether the checking will be acceptably fast.
Downloading and installing an app takes about 30 seconds on a typical Android phone over wifi.
If checking a policy delays this even further a user may become annoyed and disable AppPAL.

The policy checking procedure is at its slowest when having to delegate repeatedly;
   the depth of the delegation tree is the biggest factor for slowing the search.
Synthetic benchmarks were created to check that the checking procedure performed acceptably.
Each benchmark consisted of a chain of delegations.
The \emph{1 to 1} benchmark consists of a repeated delegation between all the principals.
In the \emph{1 to 2} benchmark each principal delegated to 2 others and in the \emph{1 to 3} benchmark each principal delegated to 3 others.
These benchmarks are reasonable as they model the slowest kinds of policies to evaluate---though worse ones could be designed by delegating even more.

@@

Throughout we have asserted that users have policies and that there is a need for policy enforcement tools.
Corporate mobile security \ac{byod} policies have started appearing and NIST have issued recommendations for writing them~\cite{Scarfone:2009vy,Souppaya:2013jf}.
In a study of 725 Android users, Lin~\etal~found four patterns that characterise user privacy preferences for apps~\cite{Sadeh:2014vq} demonstrating a refinement of Westin's privacy segmentation index~\cite{Krane:2002jo}.
Using app installation data from Carat~\cite{Oliner:2013ht,Chia:2012gz} we used AppPAL to find the apps satisfying each policy Lin~\etal~identify and measure the extent that each user was following a policy.

Lin~\etal~identified four types of user.
The \emph{Conservative} (C) users were uncomfortable allowing an app access to any personal data for any reason.
The \emph{Unconcerned} (U) users felt okay allowing access to most data for almost any reason.
The \emph{Advanced} (A) users were comfortable allowing apps access to location data but not if it was for advertising reasons.
Opinions in the largest cluster, \emph{Fencesitters} (F), varied but were broadly against collection of personal data for advertising purposes.
We wrote AppPAL policies to describe each of these behaviours as increasing sets of permissions.
These simplify the privacy policies identified by Lin~\etal~as we do not take into account the reason each app might have been collecting each permission.
Using AppPAL we could write more precise rules if we could determine why each permission was requested.
Lin~\etal~used Androguard~\cite{Desnos:ub} as well as manual analysis to determine the precise reasons for each permission~\cite{Sadeh:2014vq}.

@@

It is also interesting to discover when people install apps classified as malware.
McAfee classify malware into several categories, and provided us with a dataset of apps classified as malware and \ac{pup}.
The \emph{malicious} and \emph{trojan} categories describe traditional malware.
Other categories classify \ac{pup} such as aggressive adware.
Using AppPAL we can write policies to differentiate between different kinds of malware, characterising users who allow dangerous apps and those who install poor quality ones.
\begin{lstlisting}
"user" says "mcafee" can-say
  "malware" isKindOf(App).
"mcafee" says "trojan" can-act-as "malware".
"mcafee" says "pup" can-act-as "malware".
\end{lstlisting}
If a user is enforcing a privacy policy we might also expect them to install less malware.
We can check this by using AppPAL policies to measure the number of malwares each user had installed.

We now want to test how closely user behavior follows policies.
Installation data was taken from a partially anonymized\footnote{Users are replaced with incrementing numbers, app names are replaced with hashes to protect sensitive names.} database of installed apps captured by Carat~\cite{Oliner:2013ht}.
By calculating the hashes of known package names we see who installed what.
The initial database has over 90,000 apps and 55,000 users.
On average each Carat user installed around 90 apps each; 4,300 apps have known names.
Disregarding system apps (such as \texttt{com.android.vending}) and very common apps (Facebook, Dropbox, Whatsapp, and Twitter) we reduced the set to an average of 20 known apps per user.
To see some variations in app type, we considered only the 44,000 users who had more than 20 known apps.
Using this data, and the apps themselves taken from the Google Play Store and Android Observatory~\cite{Barrera:2012iba}, we checked which apps satisfied which policies.

\begin{figure}\centering
  \subfigure[Uptake of Lin~\etal policies.]{%
    \includegraphics[width=0.48\linewidth]{./figures/lin.eps}
    \label{fig:lin}}
  \subfigure[Uptake of malware and PUPs.]{%
    \includegraphics[width=0.48\linewidth]{./figures/malware.eps}
    \label{fig:malware}}
    \caption{Policy compliance graphs. Each histogram shows the number of users who followed a policy to a certain extent.  Users who installed no malware have been omitted from Figure~\autoref{fig:malware}.}
\end{figure}
